---
title: "Introduction to WithinHostExperiment"
author: "Cuiwei Gao"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to WithinHostExperiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Installation

```{r install, eval=FALSE}
# From Bioconductor (after acceptance):
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("WithinHostExperiment")

# Development version from GitHub:
BiocManager::install("CuiweiG/WithinHostExperiment")
```

# Introduction

`WithinHostExperiment` is a Bioconductor infrastructure package for
analysing intra-host single nucleotide variants (iSNV) from pathogen
deep sequencing data.

## The problem

Researchers performing within-host pathogen diversity analysis face a
fragmented tooling landscape. Variant callers such as iVar, LoFreq,
and Freebayes produce outputs in different formats. Quality control
relies on ad hoc scripts. There is no standard Bioconductor container
that organises within-host variant data with replicate information,
donor-recipient pairing, and QC provenance.

## What this package provides

`WithinHostExperiment` fills the infrastructure gap between upstream
variant callers and downstream analysis tools:

- A `RangedSummarizedExperiment`-based container
- Replicate-aware, uncertainty-preserving QC (flag, don't delete)
- Within-host diversity quantification
- Standardised export compatible with downstream bottleneck estimation tools

## Relationship to existing tools

| Tool | Role | Relationship |
|------|------|-------------|
| iVar, LoFreq, Freebayes | Variant calling | Upstream -- we read output |
| deepSNV | Low-frequency variant detection | Upstream -- complementary |
| ViralBottleneck | Bottleneck estimation | Downstream -- simplified frame |
| QSutils | Quasispecies diversity | Parallel -- different granularity |

## Why Bioconductor?

This package belongs in Bioconductor rather than CRAN for several
reasons:

1. **High-throughput genomic data.** Within-host pathogen diversity
   analysis operates on deep sequencing data (BAM/VCF) -- the core
   domain of Bioconductor.
2. **Shared data structures.** `WithinHostExperiment` extends
   `RangedSummarizedExperiment` and interoperates with `VRanges`,
   `GRanges`, and the `VariantAnnotation` ecosystem.
3. **Complementary to existing packages.** It provides an
   interoperability layer between upstream variant callers and
   downstream bottleneck estimation workflows through a standardised
   QC container.
4. **Reproducibility.** The flag-don't-delete QC philosophy and
   full audit logging align with Bioconductor's emphasis on
   reproducible workflows.

# Quick start

```{r quickstart, message=FALSE}
library(WithinHostExperiment)

# Import iSNV data from VCF
vcf <- system.file("extdata", "test_donor.vcf",
    package = "WithinHostExperiment")
whe <- readWithinHost(vcf,
    colData = S4Vectors::DataFrame(sample_id = "donor"),
    caller = "ivar")
whe
```

## Pipe-friendly workflow

All core functions take a `WithinHostExperiment` as their first
argument, making them compatible with R's native pipe:

```{r pipe-workflow}
library(GenomicRanges)
primers <- GRanges("test_segment", IRanges::IRanges(
    start = c(40, 880), end = c(65, 950)))

result <- whe |>
    flagISNV(ISNVFilter(minDepth = 200L, minFreq = 0.03)) |>
    flagPrimerSites(primers) |>
    passedISNV()

cat("Input:", nrow(whe), "sites ->",
    "After QC + primer masking:", nrow(result), "sites\n")
```

# Importing data

## From VCF files

Each VCF file corresponds to one sample. Sample metadata is provided
via `colData`:

```{r import-vcf}
vcf1 <- system.file("extdata", "test_donor.vcf",
    package = "WithinHostExperiment")
vcf2 <- system.file("extdata", "test_recipient.vcf",
    package = "WithinHostExperiment")

whe <- readWithinHost(
    c(vcf1, vcf2),
    colData = S4Vectors::DataFrame(
        sample_id = c("donor", "recipient"),
        host_id   = c("P1", "P2"),
        role      = c("donor", "recipient"),
        pair_id   = c("pair_1", "pair_1")
    ),
    caller = "ivar"
)
whe
```

## From iVar TSV

```{r import-tsv}
tsv <- system.file("extdata", "test_donor_ivar.tsv",
    package = "WithinHostExperiment")
whe_tsv <- readWithinHostTable(tsv, format = "ivar")
whe_tsv
```

# The WithinHostExperiment object

The container extends `RangedSummarizedExperiment`:

```{r object-structure}
# Assays: variant-level matrices (rows = sites, columns = samples)
SummarizedExperiment::assayNames(whe)

# rowRanges: genomic coordinates with ref/alt
head(SummarizedExperiment::rowRanges(whe))

# colData: sample metadata
SummarizedExperiment::colData(whe)

# Transmission pairs (from colData role + pair_id)
transmissionPairs(whe)
```

# Quality control

Accurate iSNV identification requires careful quality control.
Roder et al. (2023, *mBio*) demonstrated that choice of
variant caller and use of replicate sequencing have the most
significant impact on iSNV discovery. `WithinHostExperiment`
addresses both: it harmonises output from multiple callers
(iVar, LoFreq, Freebayes) and provides replicate-aware QC as a
first-class feature.

## Standard QC with flagISNV

`flagISNV()` marks variants that fail quality criteria. It does
**not** delete rows -- it sets `qcPass` to `FALSE`. This preserves
all data for potential re-analysis with different thresholds.

```{r qc-flag}
# Create a filter
filt <- ISNVFilter(minDepth = 200L, minFreq = 0.03)
filt

# Apply (flag, don't delete)
whe_qc <- flagISNV(whe, filt)

# Check the audit log
qcLog(whe_qc)
```

```{r qc-summary}
# Per-sample QC summary
qcSummary(whe_qc)
```

```{r qc-passed}
# Extract only QC-passed variants
clean <- passedISNV(whe_qc)
cat("Before:", nrow(whe_qc), "sites\n")
cat("After passedISNV:", nrow(clean), "sites\n")
```

## Replicate-aware QC

If your data includes technical replicates, `flagReplicateDiscordance()`
identifies sites where replicates disagree:

```{r replicate-qc}
rep1 <- system.file("extdata", "test_donor_rep1.vcf",
    package = "WithinHostExperiment")
rep2 <- system.file("extdata", "test_donor_rep2.vcf",
    package = "WithinHostExperiment")

whe_rep <- readWithinHost(
    c(rep1, rep2),
    colData = S4Vectors::DataFrame(
        sample_id       = c("rep1", "rep2"),
        replicate_group = c("donor_reps", "donor_reps"),
        replicate_id    = c("1", "2")
    ),
    caller = "ivar"
)

whe_rep <- flagReplicateDiscordance(whe_rep, freqTolerance = 0.05)
qcLog(whe_rep)
```

## Primer site filtering

For amplicon-based sequencing, variants in primer binding regions
are unreliable and should be flagged:

```{r primer-qc}
library(GenomicRanges)
primers <- GRanges("test_segment", IRanges::IRanges(
    start = c(40, 880), end = c(65, 950)))
whe_primer <- flagPrimerSites(whe_qc, primers)
qcLog(whe_primer)
```

# Diversity analysis

```{r diversity}
div <- calcDiversity(whe_qc, genomeLength = 1000L)
div
```

# Transmission analysis

## Shared variant statistics

```{r shared}
sv <- calcSharedVariants(whe_qc, pairId = "pair_1")
cat("Shared:", sv$n_shared, "\n")
cat("Donor-only:", sv$n_donor_only, "\n")
cat("Recipient-only:", sv$n_recipient_only, "\n")
```

## Quick bottleneck estimation

For a lightweight MLE estimate using the beta-binomial method:

```{r bottleneck}
nb <- quickBottleneck(whe_qc, pairId = "pair_1", maxNb = 50L)
cat("Nb MLE:", nb$Nb, "\n")
cat("95% CI: [", nb$ci[1], ",", nb$ci[2], "]\n")
```

`asViralBottleneckInput()` writes the donor and recipient frequencies
of a pair into one frame, which is the input a beta-binomial estimator
needs:

```{r vb-bridge}
vb_input <- asViralBottleneckInput(whe_qc, pairId = "pair_1")
head(vb_input)
```

# Interoperability

`WithinHostExperiment` objects can be converted to standard
Bioconductor and base R types for use with other packages.

## Convert to data.frame

```{r as-dataframe}
df <- as(whe_qc, "data.frame")
head(df)
```

## Convert to VRanges

For integration with `VariantAnnotation` workflows:

```{r as-vranges}
vr <- as(whe_qc, "VRanges")
vr
```

## Export for downstream bottleneck tools

The frame shown above under "Quick bottleneck estimation" carries the
positions and the two frequencies, which is what a beta-binomial
estimator takes. `ViralBottleneck` itself builds its objects from
per-base read counts through `CreateTransmissionObject()`, so prepare
those from the `altCount` and `refCount` assays rather than from this
frame.

# Visualisation

```{r viz-spectrum, fig.width=6, fig.height=4}
plotFrequencySpectrum(whe_qc)
```

```{r viz-pair, fig.width=5, fig.height=5}
plotPairScatter(whe_qc, pairId = "pair_1")
```

```{r viz-dashboard, fig.width=8, fig.height=6}
plotQCDashboard(whe_qc)
```

# Reporting summaries

`calcDiversity()` and `quickBottleneck()` give the figures a report on a
transmission pair usually needs:

```{r reporting-summary}
div <- calcDiversity(whe_qc, genomeLength = 1000L)
cat("Shannon entropy:", round(div$shannon[1], 2), "
")
cat("Nucleotide diversity (pi):", format(div$pi[1], digits = 3), "
")
cat("Richness:", div$richness[1], "iSNV sites
")

nb <- quickBottleneck(whe_qc, pairId = "pair_1", maxNb = 50L)
cat("Bottleneck Nb:", nb$Nb, "(95% CI:", nb$ci[1], "-", nb$ci[2], ")
")
```

The example files describe a synthetic 1,000-base segment, so these numbers
show the form of the output and are not evidence that the estimators are
accurate. `validateDiversity()`, used below, compares an estimate with the
package's own plausibility range for a pathogen: a sanity check, not a
validation.

# Case study: SARS-CoV-2 transmission pair

This section demonstrates the complete workflow on real SARS-CoV-2
iSNV data from Bendall et al. (2023) *Nature Communications* 14:272.
We analyse household HH6 -- an epidemiologically confirmed
donor-recipient pair.

```{r case-study-import}
donor_file <- system.file("extdata", "MHM1143_rep1.tsv",
    package = "WithinHostExperiment")
recip_file <- system.file("extdata", "MHM1184_rep1.tsv",
    package = "WithinHostExperiment")

whe_sars2 <- readWithinHostTable(
    c(donor_file, recip_file),
    format = "ivar",
    colData = S4Vectors::DataFrame(
        sample_id = c("MHM1143", "MHM1184"),
        role = c("donor", "recipient"),
        pair_id = c("HH6", "HH6")
    )
)
whe_sars2
```

## QC and diversity

```{r case-study-qc}
whe_sars2 <- flagISNV(whe_sars2,
    ISNVFilter(minDepth = 100L, minFreq = 0.03))
qcSummary(whe_sars2)

div_sars2 <- calcDiversity(whe_sars2, genomeLength = 29903L)
div_sars2
```

## Transmission analysis

```{r case-study-transmission}
sv <- calcSharedVariants(whe_sars2, pairId = "HH6")
cat("Shared:", sv$n_shared, "/ Donor-only:", sv$n_donor_only,
    "/ Recipient-only:", sv$n_recipient_only, "\n")
```

# Other pathogens

To show that the workflow is not specific to SARS-CoV-2, we run it on
small synthetic example files for three further pathogens and check the
diversity estimates against the plausibility ranges of
`validateDiversity()`. The files are illustrative fixtures, not
measurements, and the ranges are the package's own heuristics, so these
sections demonstrate the interface rather than validate the estimates.

## Influenza A (H3N2 HA segment)

```{r multipathogen-flu}
flu_file <- system.file("extdata", "test_influenza_donor.tsv",
    package = "WithinHostExperiment")
whe_flu <- readWithinHostTable(flu_file, format = "ivar",
    colData = S4Vectors::DataFrame(sample_id = "flu_donor"))
whe_flu <- flagISNV(whe_flu, adaptiveFilter("illumina"))
div_flu <- calcDiversity(whe_flu, genomeLength = 1701L,
    indices = c("pi", "shannon"))
cat("Influenza pi:", format(div_flu$pi[1], digits = 3), "\n")
cat("Influenza Shannon:", round(div_flu$shannon[1], 2), "\n")
validateDiversity(div_flu, pathogen = "influenza")
```

## HIV-1 (pol region)

```{r multipathogen-hiv}
hiv_file <- system.file("extdata", "test_hiv_sample.tsv",
    package = "WithinHostExperiment")
whe_hiv <- readWithinHostTable(hiv_file, format = "ivar",
    colData = S4Vectors::DataFrame(sample_id = "hiv_sample"))
whe_hiv <- flagISNV(whe_hiv, adaptiveFilter("illumina"))
div_hiv <- calcDiversity(whe_hiv, genomeLength = 1000L,
    indices = c("pi", "shannon"))
cat("HIV pi:", format(div_hiv$pi[1], digits = 3), "\n")
cat("HIV Shannon:", round(div_hiv$shannon[1], 2), "\n")
validateDiversity(div_hiv, pathogen = "hiv")
```

## Mycobacterium tuberculosis

```{r multipathogen-tb}
tb_file <- system.file("extdata", "test_tb_sample.tsv",
    package = "WithinHostExperiment")
whe_tb <- readWithinHostTable(tb_file, format = "ivar",
    colData = S4Vectors::DataFrame(sample_id = "tb_sample"))
whe_tb <- flagISNV(whe_tb, adaptiveFilter("illumina"))
div_tb <- calcDiversity(whe_tb, genomeLength = 4411532L,
    indices = c("pi", "shannon"))
cat("TB pi:", format(div_tb$pi[1], digits = 3), "\n")
cat("TB Shannon:", round(div_tb$shannon[1], 2), "\n")
validateDiversity(div_tb, pathogen = "tb")
```

## Neutrality tests across pathogens

```{r multipathogen-neutrality}
nt_flu <- calcNeutralityTests(whe_flu, genomeLength = 1701L)
nt_hiv <- calcNeutralityTests(whe_hiv, genomeLength = 1000L)
nt_tb  <- calcNeutralityTests(whe_tb,  genomeLength = 4411532L)
results <- data.frame(
    pathogen = c("Influenza", "HIV-1", "M. tuberculosis"),
    S = c(nrow(whe_flu), nrow(whe_hiv), nrow(whe_tb)),
    tajimaD = c(nt_flu$tajimaD, nt_hiv$tajimaD, nt_tb$tajimaD),
    fusFs = c(nt_flu$fusFs, nt_hiv$fusFs, nt_tb$fusFs)
)
results
```

Negative Tajima's D and Fu's Fs values are expected for within-host
pathogen populations, consistent with purifying selection and/or
recent population expansion from a transmission bottleneck.

# Advanced features

This section demonstrates additional analytical modules available
in `WithinHostExperiment`. Each subsection provides a minimal
working example.

## Neutrality tests

Tajima's D and Fu's Fs detect departures from neutral evolution.
The `calcNeutralityTests()` method caps the effective sample size
at *n* = 100 to avoid deep-sequencing bias (see `?tajimaD`).

```{r neutrality}
nt <- calcNeutralityTests(whe_qc, genomeLength = 1000L)
nt
```

## GFF3 gene annotation

Annotate variant sites with gene names from a GFF3 file. This
automates the positional overlap needed for `dndsWithinHost()`:

```{r gff-annotate}
gff_file <- tempfile(fileext = ".gff3")
writeLines(c(
    "##gff-version 3",
    "test_segment\t.\tgene\t1\t500\t.\t+\t.\tName=geneA",
    "test_segment\t.\tgene\t501\t1000\t.\t+\t.\tName=geneB"
), gff_file)
whe_ann <- annotateFromGFF(whe_qc, gff_file)
table(S4Vectors::mcols(
    SummarizedExperiment::rowRanges(whe_ann))$GFF_FEATURE)
```

For codon-aware annotation (amino acid changes), use
`annotateCodonChange()` with a GFF3 and reference FASTA:

```{r codon-annotate}
ref_fa <- system.file("extdata", "test_ref.fa",
    package = "WithinHostExperiment")
if (file.exists(ref_fa)) {
    gff_cds <- tempfile(fileext = ".gff3")
    writeLines(c("##gff-version 3",
        "test_segment\t.\tCDS\t1\t999\t.\t+\t0\tgene=testGene"),
        gff_cds)
    whe_codon <- annotateCodonChange(whe_qc, gff_cds, ref_fa)
    mc <- S4Vectors::mcols(
        SummarizedExperiment::rowRanges(whe_codon))
    table(mc$AA_CLASS)
}
```

For complex genomes with overlapping reading frames or
non-standard codon usage, consider SnpEff or ANNOVAR upstream.

## Selection analysis

### Within-host dN/dS

`dndsWithinHost()` counts synonymous and nonsynonymous iSNVs, divides
each by the synonymous and nonsynonymous sites of the coding
sequences (Nei and Gojobori 1986) and applies the Jukes-Cantor
correction. It uses the codon annotation from `annotateCodonChange()`
together with the same GFF3 and reference FASTA:

```{r dnds}
if (file.exists(ref_fa)) {
    dnds <- dndsWithinHost(whe_codon, gff = gff_cds, refFasta = ref_fa)
    dnds[, c("sample_id", "nS", "nN", "S_sites", "N_sites", "dNdS")]
}
```

### Selection coefficients from frequency trajectories

For longitudinal data, estimate the selection coefficient from
the logit-transformed frequency change:

```{r selection}
freqs <- c(0.05, 0.10, 0.25, 0.50)
s_hat <- estimateSelectionCoefficient(freqs,
    times = c(0, 3, 7, 14))
s_hat
```

## Multi-caller consensus

When data from multiple variant callers are available, create a
consensus call set:

```{r consensus}
gr1 <- GenomicRanges::GRanges("seg",
    IRanges::IRanges(c(100, 200, 300), width = 1))
S4Vectors::mcols(gr1)$ref <- c("A", "C", "G")
S4Vectors::mcols(gr1)$alt <- c("T", "G", "A")
whe1 <- WithinHostExperiment(
    assays = list(altFreq = matrix(c(0.10, 0.20, 0.30), ncol = 1)),
    rowRanges = gr1,
    colData = S4Vectors::DataFrame(sample_id = "S1"))

gr2 <- GenomicRanges::GRanges("seg",
    IRanges::IRanges(c(100, 200, 400), width = 1))
S4Vectors::mcols(gr2)$ref <- c("A", "C", "T")
S4Vectors::mcols(gr2)$alt <- c("T", "G", "C")
whe2 <- WithinHostExperiment(
    assays = list(altFreq = matrix(c(0.12, 0.18, 0.15), ncol = 1)),
    rowRanges = gr2,
    colData = S4Vectors::DataFrame(sample_id = "S1"))

cons <- consensusISNV(list(ivar = whe1, lofreq = whe2))
cons

## Pairwise caller concordance
calcCallerConcordance(list(ivar = whe1, lofreq = whe2))
```

## Exact bottleneck estimation

`exactBottleneck()` offers three estimators beyond the approximate
`quickBottleneck()`: `"exact_bb"`, `"wright_fisher"` and
`"presence_absence"`. The presence-absence estimator is shown here.

```{r exact-bottleneck}
nb_pa <- exactBottleneck(whe_qc, pairId = "pair_1",
    maxNb = 50L, method = "presence_absence")
cat("Presence-absence Nb:", nb_pa$Nb,
    "(95% CI:", nb_pa$ci[1], "-", nb_pa$ci[2], ")\n")
```

## Longitudinal frequency tracking

Track variant frequencies across time points within a host:

```{r longitudinal}
gr_l <- GenomicRanges::GRanges("seg1",
    IRanges::IRanges(c(100, 200), width = 1))
S4Vectors::mcols(gr_l)$ref <- c("A", "C")
S4Vectors::mcols(gr_l)$alt <- c("T", "G")
freq_l <- matrix(c(0.02, 0.10, 0.12, 0.15),
    nrow = 2, ncol = 2)
whe_l <- WithinHostExperiment(
    assays = list(altFreq = freq_l),
    rowRanges = gr_l,
    colData = S4Vectors::DataFrame(
        sample_id = c("t1", "t2"),
        host_id = c("H1", "H1"),
        timepoint = c(1, 2)))

## Track frequencies
trackFrequency(whe_l)

## Detect emerging variants
detectEmergingVariants(whe_l)
```

## SCI strand bias filtering

The SCI score normalises strand bias by the global strand coverage
ratio, which suits amplicon data with unequal strand coverage:

```{r sci-strand}
sci <- sciStrandBias(
    fwd_alt   = c(50, 2, 30),
    rev_alt   = c(45, 40, 28),
    fwd_total = c(500, 500, 500),
    rev_total = c(480, 480, 480))
sci
```

## Adaptive technology-specific filtering

Pre-configured filter presets for different sequencing platforms:

```{r adaptive}
adaptiveFilter("illumina")
adaptiveFilter("ont")
adaptiveFilter("pacbio_hifi")
```

## Diversity modelling and statistical utilities

Fit linear models to diversity data and correct for multiple
testing:

```{r modeling}
div_df <- data.frame(
    shannon = c(1.2, 1.5, 0.8, 1.1, 1.9, 0.7),
    timepoint = c(1, 2, 3, 1, 2, 3),
    group = c("A", "A", "A", "B", "B", "B"))
res <- fitDiversityModel(div_df, shannon ~ timepoint,
    group_col = "group")
res$summary_df

## Multiple testing correction
raw_p <- c(0.001, 0.04, 0.03, 0.5)
multitestCorrection(raw_p)
```

## Plausibility check

Cross-check observed diversity against the package's plausibility
range for a given pathogen:

```{r validate}
div_check <- S4Vectors::DataFrame(pi = 1e-4, shannon = 5.0)
validateDiversity(div_check, pathogen = "sars_cov_2")
```

# Scope and limitations

This package operates at the level of individual single nucleotide
variants (iSNVs). It does **not** perform:

- **Haplotype phasing or quasispecies reconstruction**, which require
  read-level (BAM) analysis outside the scope of
  `RangedSummarizedExperiment`. For haplotype-aware analysis, consider
  CliqueSNV (Knyazev et al. 2021 *Nucleic Acids Res* 49:e102), ShoRAH
  (Zagordi et al. 2011 *BMC Bioinformatics* 12:119), or similar
  tools. Their output can be imported into `WithinHostExperiment`
  for downstream diversity and QC analysis.

- **Co-infection vs diversity distinction**, which requires
  phylogenetic analysis of full haplotypes. The multi-caller
  consensus and replicate-concordance modules reduce artefacts
  but do not resolve this.

- **Stochastic selection coefficient estimation**. The
  `estimateSelectionCoefficient()` function uses a deterministic
  logit-frequency model. For drift-aware inference, consider
  dedicated Wright-Fisher or diffusion-based frameworks.

# Session info

```{r session}
sessionInfo()
```
